---
title: "Moonlight2: An automatized workflow to study mechanistic indicators for driver gene prediction"
author: "Astrid Saksager^+^ ,
        Mona Nourbakhsh^+^,
        Nikola Tom, 
        Xi Steven Chen,
        Antonio Colaprico,
        Catharina Olsen,
        Matteo Tiberti,
        Elena Papaleo"
subtitle: ^+^ These authors contributed equally to the paper as first authors.

date: "`r Sys.Date()`"

output: 
    BiocStyle::html_document:
        toc: true
        number_sections: false
        toc_depth: 2
        highlight: haddock


references:
- id: ref1
  title: Orchestrating high-throughput genomic analysis with Bioconductor
  author: 
  - family: Huber, Wolfgang and Carey, Vincent J and Gentleman, Robert and Anders, Simon and Carlson, Marc and Carvalho, Benilton S and Bravo, Hector Corrada and Davis, Sean and Gatto, Laurent and Girke, Thomas and others
    given:
  journal: Nature methods
  volume: 12
  number: 2
  pages: 115-121
  issued:
    year: 2015

- id: ref2
  title: GC-content normalization for RNA-Seq data
  author: 
  - family: Risso, Davide and Schwartz, Katja and Sherlock, Gavin and Dudoit, Sandrine
    given:
  journal: BMC bioinformatics
  volume: 12
  number: 1
  pages: 480
  issued:
    year: 2011 
    
- id: ref3
  title: Evaluation of statistical methods for normalization and differential expression in mRNA-Seq experiments
  author: 
  - family: Bullard, James H and Purdom, Elizabeth and Hansen, Kasper D and Dudoit, Sandrine
    given:
  journal: BMC bioinformatics
  volume: 11
  number: 1
  pages: 94
  issued:
    year: 2010 

- id: ref4
  title: Inferring regulatory element landscapes and transcription factor networks from cancer methylomes
  author: 
  - family: Yao, L., Shen, H., Laird, P. W., Farnham, P. J., & Berman, B. P. 
    given:
  journal: Genome biology
  volume: 16
  number: 1
  pages: 105
  issued:
    year: 2015 

- id: ref5
  title: Evaluation of statistical methods for normalization and differential expression in mRNA-Seq experiments
  author: 
  - family:  James H Bullard, Elizabeth Purdom, Kasper D Hansen and Sandrine Dudoit	
    given:
  journal: BMC Bioinformatics
  volume: 11
  number: 1
  pages: 94
  issued:
    year: 2010 

- id: ref6
  title: GC-content normalization for RNA-Seq data
  author: 
  - family: Risso, D., Schwartz, K., Sherlock, G., & Dudoit, S. 
    given:
  journal: BMC Bioinformatics
  volume: 12
  number: 1
  pages: 480
  issued:
    year: 2011 

- id: ref7
  title: Identification of a CpG island methylator phenotype that defines a distinct subgroup of glioma
  author: 
  - family: Noushmehr, H., Weisenberger, D.J., Diefes, K., Phillips, H.S., Pujara, K., Berman, B.P., Pan, F., Pelloski, C.E., Sulman, E.P., Bhat, K.P. et al.
    given:
  journal: Cancer cell
  volume: 17
  number: 5
  pages: 510-522
  issued:
    year: 2010

- id: ref8
  title: Molecular Profiling Reveals Biologically Discrete Subsets and Pathways of Progression in Diffuse Glioma
  author: 
  - family: Ceccarelli, Michele and Barthel, Floris P and Malta, Tathiane M and Sabedot, Thais S and Salama, Sofie R and Murray, Bradley A and Morozova, Olena and Newton, Yulia and Radenbaugh, Amie and Pagnotta, Stefano M and others
    given:
  journal: Cell
  URL: "http://doi.org/10.1016/j.cell.2015.12.028"
  DOI: "10.1016/j.cell.2015.12.028"
  volume: 164
  number: 3
  pages: 550-563
  issued:
    year: 2016


- id: ref9
  title: Comprehensive molecular profiling of lung adenocarcinoma
  author: 
  - family: Cancer Genome Atlas Research Network and others
    given:
  journal: Nature
  URL: "http://doi.org/10.1038/nature13385"
  DOI: "10.1038/nature13385"
  volume: 511
  number: 7511
  pages: 543-550
  issued:
    year: 2014


- id: ref10
  title: Comprehensive molecular characterization of gastric adenocarcinoma
  author: 
  - family: Cancer Genome Atlas Research Network and others
    given:
  journal: Nature
  URL: "http://doi.org/10.1038/nature13480"
  DOI: "10.1038/nature13480"
  issued:
    year: 2014

- id: ref11
  title: Comprehensive molecular portraits of human breast tumours
  author: 
  - family: Cancer Genome Atlas Research Network and others
    given:
  journal: Nature
  URL: "http://doi.org/10.1038/nature11412"
  DOI: "10.1038/nature11412"
  volume: 490
  number: 7418
  pages: 61-70
  issued:
    year: 2012
  
- id: ref12
  title: Comprehensive molecular characterization of human colon and rectal cancer
  author: 
  - family: Cancer Genome Atlas Research Network and others
    given:
  journal: Nature
  URL: "http://doi.org/10.1038/nature11252"
  DOI: "10.1038/nature11252"
  volume: 487
  number: 7407
  pages: 330-337
  issued:
    year: 2012    

- id: ref13
  title: Genomic classification of cutaneous melanoma
  author: 
  - family: Cancer Genome Atlas Research Network and others
    given:
  journal: Cell
  URL: "http://doi.org/10.1016/j.cell.2015.05.044"
  DOI: "10.1016/j.cell.2015.05.044"
  volume: 161
  number: 7
  pages: 1681-1696
  issued:
    year: 2015    

- id: ref14
  title: Comprehensive genomic characterization of head and neck squamous cell carcinomas
  author: 
  - family: Cancer Genome Atlas Research Network and others
    given:
  journal: Nature
  URL: "http://doi.org/10.1038/nature14129"
  DOI: "10.1038/nature14129"
  volume: 517
  number: 7536
  pages: 576-582
  issued:
    year: 2015    

- id: ref15
  title: The somatic genomic landscape of chromophobe renal cell carcinoma
  author: 
  - family: Davis, Caleb F and Ricketts, Christopher J and Wang, Min and Yang, Lixing and Cherniack, Andrew D and Shen, Hui and Buhay, Christian and Kang, Hyojin and Kim, Sang Cheol and Fahey, Catherine C and others
    given:
  journal: Cancer Cell
  URL: "http://doi.org/10.1016/j.ccr.2014.07.014"
  DOI: "10.1016/j.ccr.2014.07.014"
  volume: 26
  number: 3
  pages: 319-330
  issued:
    year: 2014    


- id: ref16
  title: Comprehensive genomic characterization of squamous cell lung cancers
  author: 
  - family: Cancer Genome Atlas Research Network and others
    given:
  journal: Nature
  URL: "http://doi.org/10.1038/nature11404"
  DOI: "10.1038/nature11404"
  volume: 489
  number: 7417
  pages: 519-525
  issued:
    year: 2012   

- id: ref17
  title: Integrated genomic characterization of endometrial carcinoma
  author: 
  - family: Cancer Genome Atlas Research Network and others
    given:
  journal: Nature
  URL: "http://doi.org/10.1038/nature12113"
  DOI: "10.1038/nature12113"
  volume: 497
  number: 7447
  pages: 67-73
  issued:
    year: 2013   

- id: ref18
  title: Integrated genomic characterization of papillary thyroid carcinoma
  author: 
  - family: Cancer Genome Atlas Research Network and others
    given:
  journal: Cell
  URL: "http://doi.org/10.1016/j.cell.2014.09.050"
  DOI: "10.1016/j.cell.2014.09.050"
  volume: 159
  number: 3
  pages: 676-690
  issued:
    year: 2014   

- id: ref19
  title: The molecular taxonomy of primary prostate cancer
  author: 
  - family: Cancer Genome Atlas Research Network and others
    given:
  journal: Cell
  URL: "http://doi.org/10.1016/j.cell.2015.10.025"
  DOI: "10.1016/j.cell.2015.10.025"
  volume: 163
  number: 4
  pages: 1011-1025
  issued:
    year: 2015   
    

- id: ref20
  title: Comprehensive Molecular Characterization of Papillary Renal-Cell Carcinoma
  author: 
  - family: Linehan, W Marston and Spellman, Paul T and Ricketts, Christopher J and Creighton, Chad J and Fei, Suzanne S and Davis, Caleb and Wheeler, David A and Murray, Bradley A and Schmidt, Laura and Vocke, Cathy D and others
    given:
  journal: NEW ENGLAND JOURNAL OF MEDICINE
  URL: "http://doi.org/10.1056/NEJMoa1505917"
  DOI: "10.1056/NEJMoa1505917"
  volume: 374
  number: 2
  pages: 135-145
  issued:
    year: 2016    
    
- id: ref21
  title: Comprehensive molecular characterization of clear cell renal cell carcinoma
  author: 
  - family: Cancer Genome Atlas Research Network and others
    given:
  journal: Nature
  URL: "http://doi.org/10.1038/nature12222"
  DOI: "10.1038/nature12222"
  volume: 499
  number: 7456
  pages: 43-49
  issued:
    year: 2013        
          
- id: ref22
  title: Comprehensive Pan-Genomic Characterization of Adrenocortical Carcinoma
  author: 
  - family: Cancer Genome Atlas Research Network and others
    given:
  journal: Cancer Cell
  URL: "http://dx.doi.org/10.1016/j.ccell.2016.04.002"
  DOI: "10.1016/j.ccell.2016.04.002"
  volume: 29
  pages: 43-49
  issued:
    year: 2016 

- id: ref23
  title: Complex heatmaps reveal patterns and correlations in multidimensional genomic data
  author: 
  - family: Gu, Zuguang and Eils, Roland and Schlesner, Matthias
  given:
  journal: Bioinformatics
  URL: "http://dx.doi.org/10.1016/j.ccell.2016.04.002"
  DOI: "10.1016/j.ccell.2016.04.002"
  pages: "btw313"
  issued:
   year: 2016 

- id: ref24
  title: "TCGA Workflow: Analyze cancer genomics and epigenomics data using Bioconductor packages"
  author: 
  - family:  Silva, TC and Colaprico, A and Olsen, C and D'Angelo, F and Bontempi, G and Ceccarelli, M and Noushmehr, H
  given:
  journal: F1000Research
  URL: "http://dx.doi.org/10.12688/f1000research.8923.1"
  DOI: "10.12688/f1000research.8923.1"
  volume: 5
  number: 1542
  issued:
   year: 2016 

- id: ref25
  title: "TCGAbiolinks: an R/Bioconductor package for integrative analysis of TCGA data"
  author: 
  - family:  Colaprico, Antonio and Silva, Tiago C. and Olsen, Catharina and Garofano, Luciano and Cava, Claudia and Garolini, Davide and Sabedot, Thais S. and Malta, Tathiane M. and Pagnotta, Stefano M. and Castiglioni, Isabella and Ceccarelli, Michele and Bontempi, Gianluca and Noushmehr, Houtan
  given:
  journal: Nucleic Acids Research
  URL: "http://dx.doi.org/10.1093/nar/gkv1507"
  DOI: "10.1093/nar/gkv1507"
  volume: 44
  number: 8
  pages: e71
  issued:
   year: 2016 

- id: ref26
  title: "Interpreting pathways to discover cancer driver genes with Moonlight"
  author: 
  - family:  Colaprico, Antonio and Olsen, Catharina and Bailey, Matthew H. and Odom, Gabriel J. and Terkelsen, Thilde and Silva, Tiago C. and Olsen, André V. and Cantini, Laura and Zinovyev, Andrei and Barillot, Emmanuel and Noushmehr, Houtan and Bertoli, Gloria and Castiglioni, Isabella and Cava, Claudia and Bontempi, Gianluca and Chen, Xi Steven and Papaleo, Elena
  given:
  journal: Nature Communications
  URL: "https://doi.org/10.1038/s41467-019-13803-0"
  DOI: "10.1038/s41467-019-13803-0"
  volume: 11
  number: 69
  issued:
   year: 2020

- id: ref27
  title: "An Automatized Workflow to Study Mechanistic Indicators for Driver Gene Prediction with Moonlight"
  author: 
  - family:  Saksager, Astrid and Nourbakhsh, Mona and Tom, Nikola and Chen, Xi Steven and Colaprico, Antonio and Olsen, Catharina and Tiberti, Matteo and Papaleo, Elena
  given:
  journal: bioRxiv
  URL: "https://www.biorxiv.org/content/10.1101/2022.11.18.517066v1"
  DOI: "10.1101/2022.11.18.517066"
  issued:
   year: 2022

vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---
  
```{r setup, include=FALSE}
knitr::opts_chunk$set(dpi = 300)
knitr::opts_chunk$set(cache=FALSE)
```

```{r, echo = FALSE,hide=TRUE, message=FALSE,warning=FALSE}
library(dbplyr)
devtools::load_all(".")
```

# Abstract

In order to make light of cancer development, it is crucial to understand which genes play 
a role in the mechanisms linked to this disease and moreover which role that is. Commonly 
biological processes such as proliferation and apoptosis have been linked to cancer progression. 
We have developed the Moonlight framework that allows for prediction of cancer driver genes 
through multi-omics data integration. Based on expression data we perform functional enrichment 
analysis, infer gene regulatory networks and upstream regulator analysis to score the importance 
of well-known biological processes with respect to the studied cancer. We then use these scores 
to predict oncogenic mediators with two specific roles: genes that potentially act as tumor 
suppressor genes (TSGs) and genes that potentially act as oncogenes (OCGs). This constitutes 
Moonlight's primary layer. As gene expression data alone does not explain the cancer phenotypes, 
a second layer of evidence is needed. We have automated the integration of a secondary mutational 
layer that predicts driver mutations in the oncogenic mediators and thereby allows for the 
prediction of cancer driver genes using the driver mutation prediction tool CScape-somatic. 
These new functionalities are provided in the updated version of Moonlight, namely Moonlight2. 
Overall, this methodology not only allows us to identify genes with dual role (TSG in one cancer 
type and OCG in another) but also to elucidate the underlying biological processes. 

# Introduction 

Cancer development is influenced by mutations in two distinctly different categories of genes, 
known as  tumor suppressor genes (TSG) and oncogenes (OCG). The occurrence of mutations in genes 
of the first category leads to faster cell proliferation while mutations in genes of second 
category increases or changes their function.
In 2020, we developed the Moonlight framework that allows for prediction of cancer driver genes 
[@ref26]. Here, gene expression data are integrated together with 
biological processes and gene regulatory networks to score the importance of well-known biological 
processes with respect to the studied cancer. These scores are used to predict oncogenic mediators: 
putative TSGs and putative OCGs. As gene expression data alone is not enough to explain the 
deregulation of the genes, a second layer of evidence is needed. For this reason, we automated the 
integration of a secondary mutational layer which predicts driver mutations and passenger mutations 
in the oncogenic mediators. These new functionalities are released in the updated version of 
Moonlight to produce Moonlight2R. The prediction of the driver mutations are carried out using 
the CScape-somatic driver mutation prediction tool. Moreover, the new functionalities estimate 
the potential effect of a mutation on the transcriptional, translational, or protein 
structure/function level. Those oncogenic mediators with at least one driver mutation are retained
as the final set of driver genes [@ref27]. 

# Moonlight's pipeline

Moonlight's pipeline is shown below:
```{r, fig.width=6, fig.height=4, echo = FALSE, fig.align="center",hide=TRUE, message=FALSE,warning=FALSE}
library(png)
library(grid)
img <- readPNG("Moonlight2_pipeline.png")
grid.raster(img)
```

# Moonlight's proposed workflow

The proposed pipeline consists of following eight steps:

1. The input to Moonlight is a set of differentially expressed genes between two biological conditions 
such as cancer and healthy samples or two cancer subtyoes. Besides differentially expressed genes,
gene expression and mutation data are also needed. 
2. **FEA** Functional Enrichment Analysis, using Fisher's test, to identify gene sets (with biological 
functions linked to cancer) significantly enriched by regulated genes (RG).
3. **GRN** Gene Regulatory Network inferred between each single DEG (sDEG) and all genes by means of 
mutual information, obtaining for each DEG a list of RG.
4. **URA** Upstream Regulator Analysis for DEGs in each enriched gene set, we applied z-score being 
the ratio between the sum of all predicted effects for all the gene involved in the specific function 
and the square-root of the number of all genes.
5. **PRA** Pattern Recognition Analysis identifies putative TSGs (down) and OCGs (up). We either use 
user defined biological processes or random forests.
6. **DMA** Driver Mutation Analysis identifies driver mutations in the putative TSGs and OCGs through 
the driver mutation prediction tool, CScape-somatic.
7. We applied the above procedure to multiple cancer types to obtain cancer-specific lists of TSGs 
and OCGs. We compared the lists for each cancer: if a sDEG was TSG in a cancer and OCG in another we 
defined it as dual-role TSG-OCG. Otherwise if we found a sDEG defined as OCG or TSG only in one tissue 
we defined it as tissue specific biomarker.
8. We use the COSMIC database to define a list of gold standard TSG and OCGs to assess the accuracy 
of the proposed method.


# Installation

To install Moonlight2R use the code below.
Moonlight2R requires R version >= 4.2.

First, install `devtools` or if you already have it installed, load it.
```{r, eval = FALSE}
install.packages("devtools")
library(devtools)
```

Install Moonlight2R from GitHub:
```{r, eval = FALSE}
devtools::install_github(repo = "ELELAB/Moonlight2R")
```

To install Moonlight2R with its accompanying vignette use the code below.  

First, install the BiocStyle Bioconductor package.

```{r, eval = FALSE}
if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install("BiocStyle")
```
Then install Moonlight2R with its accompanying vignette.  

```{r, eval = FALSE}
devtools::install_github(repo = "ELELAB/Moonlight2R", build_vignettes = TRUE)
```

You can view the vignette in the following way.  

```{r, eval = FALSE}
vignette( "Moonlight2R", package="Moonlight2R")
```


# Load libraries
```{r, eval = FALSE}
library(Moonlight2R)
```


# `Obtain Input`

The input to Moonlight is a set of differentially expressed genes and gene 
expression and mutation data are also needed. Gene expression data, mutation 
data and differentially expressed genes can for example be obtained from TCGA 
using the R package TCGAbiolinks. Help documents on how to use TCGAbiolinks
are available [here](https://bioconductor.org/packages/release/bioc/vignettes/TCGAbiolinks/inst/doc/index.html).
To find other examples of usage of TCGAbiolinks on TCGA cancer types see our
[GitHub repository](https://github.com/ELELAB/LUAD_LUSC_TCGA_comparison).
Example data of the input (differentially expressed genes, gene expression data, 
and mutation data) are stored in the Moonlight2R package:

```{r, eval = TRUE}
data(DEGsmatrix)
data(dataFilt)
data(dataMAF)
```


# `Download`: Get GEO data

You can search GEO data using the `getDataGEO` function.

GEO_TCGAtab: a 18x12 matrix that provides the GEO data set we matched to one of the 18 given TCGA cancer types 

```{r, eval = TRUE, echo = TRUE}
knitr::kable(GEO_TCGAtab, digits = 2, 
             caption = "Table with GEO data set matched to one 
             of the 18 given TCGA cancer types ",
             row.names = TRUE)
```



## `getDataGEO`: Search by cancer type and data type [Gene Expression]
The user can query and download the cancer types supported by GEO, using the function `getDataGEO`:

```{r, eval = FALSE , echo = TRUE, results='hide', warning = FALSE, message = FALSE}
dataFilt_GEO <- getDataGEO(GEOobject = "GSE20347",platform = "GPL571")
```

```{r, eval = FALSE, echo = TRUE, results='hide', warning = FALSE, message = FALSE}
dataFilt_GEO <- getDataGEO(TCGAtumor = "ESCA")
```


## `FEA`: Functional Enrichment Analysis
The user can perform a functional enrichment analysis using the function `FEA`. 
For each DEG in the gene set a z-score is calculated. This score indicates how the genes act in the gene set.

```{r, eval = TRUE, echo = TRUE, results='hide'}
data(DEGsmatrix)
dataFEA <- FEA(DEGsmatrix = DEGsmatrix)
```

The output can be visualized with a FEA plot. 

## `FEAplot`: Functional Enrichment Analysis Plot
The user can plot the result of a functional enrichment analysis using the function `plotFEA`. 
A negative z-score indicates that the process' activity is decreased.  A positive z-score 
indicates that the process' activity is increased. 

```{r, eval = TRUE, echo = TRUE, message=FALSE, results='hide', warning=FALSE}
plotFEA(dataFEA = dataFEA, additionalFilename = "_exampleVignette", height = 20, width = 10)
```

The figure generated by the above code is shown below:
```{r, fig.width=6, fig.height=4, echo = FALSE, fig.align="center",hide=TRUE, message=FALSE,warning=FALSE}
img <- readPNG("FEAplot.png")
grid.raster(img)
```


## `GRN`: Gene Regulatory Network
The user can perform a gene regulatory network analysis using the function `GRN` which infers 
the network using the parmigene package.

```{r, eval = FALSE}
data(DEGsmatrix)
data(dataFilt)
dataGRN <- GRN(TFs = rownames(DEGsmatrix), normCounts = dataFilt,
	               nGenesPerm = 2000,kNearest = 3,nBoot = 400)
```


## `URA`: Upstream Regulator Analysis
The user can perform upstream regulator analysis using the function `URA`. This function is 
applied to each DEG in the enriched gene set and its neighbors in the GRN.

```{r, eval = TRUE, echo = TRUE, results='hide'}
data(dataGRN)
data(DEGsmatrix)

dataFEA <- FEA(DEGsmatrix = DEGsmatrix)

BPselected <- dataFEA$Diseases.or.Functions.Annotation[1:5]
dataURA <- URA(dataGRN = dataGRN,
               DEGsmatrix = DEGsmatrix,
               BPname = BPselected, 
               nCores=1)
```

## `PRA`: Pattern Regognition Analysis
The user can retrieve TSG/OCG candidates using either selected biological processes or a 
random forest classifier trained on known COSMIC OCGs/TSGs.  

```{r, eval = TRUE}
data(dataURA)
dataPRA <- PRA(dataURA = dataURA,
               BPname = c("apoptosis","proliferation of cells"),
               thres.role = 0)
```

## `DMA`: Driver Mutation Analysis
The user can identify driver mutations with `DMA` in the oncogenic mediators established by `PRA`. 
The passenger or driver status is estimated with CScape-somatic.  
This function will further generate three files: DEG_Mutations_Annotations.rda, 
Oncogenic_mediators_mutation_summary.rda and cscape_somatic_output.rda. These will be placed 
in the specified results-folder. 
The user needs to download two CScape-somatic files in order to run DMA named css_coding.vcf.gz
and css_noncoding.vcf.gz, respectively. These two files can be downloaded at 
http://cscape-somatic.biocompute.org.uk/#download. The corresponding .tbi files (css_coding.vcf.gz.tbi
and css_noncoding.vcf.gz.tbi) must also be downloaded and be placed in the same folder.


```{r, eval = FALSE}
data(dataPRA)
data(dataMAF)
data(DEGsmatrix)
dataDMA <- DMA(dataMAF = dataMAF,
               dataDEGs = DEGsmatrix, 
               dataPRA = dataPRA,
               results_folder = "DMAresults",
               coding_file = "css_coding.vcf.gz",
               noncoding_file = "css_noncoding.vcf.gz")
```

## `Level of consequence`: Effect of mutations on three different levels
The user can investigate the predicted effect of different mutation types on the transcriptional 
level through the table LOC_transcription:

```{r}
knitr::kable(LOC_transcription)
```

The user can investigate the predicted effect of different mutation types on the translational 
level through the table LOC_translation:

```{r}
knitr::kable(LOC_translation)
```

The user can investigate the predicted effect of different mutation types on the protein level 
through the table LOC_protein:
```{r}
knitr::kable(LOC_protein)
```


## `plotNetworkHive`: GRN hive visualization taking into account COSMIC cancer genes

In the following plot the nodes are separated into three groups: known tumor suppressor genes (yellow), 
known oncogenes (green) and the rest (gray).

```{r, eval = TRUE, echo = TRUE, results='hide', warning = FALSE, message = FALSE}
data(knownDriverGenes)
data(dataGRN)
plotNetworkHive(dataGRN, knownDriverGenes, 0.55)
```


## `plotDMA`: Heatmap of the driver/passenger status of mutations in TSGs/OCGs

In the following plot the driver genes with driver mutations are shown.

```{r, eval = FALSE,  warning = FALSE, message = FALSE, include=TRUE}
data(dataDMA)
data(DEG_Mutations_Annotations)
data(Oncogenic_mediators_mutation_summary)
plotDMA(DEG_Mutations_Annotations, 
        Oncogenic_mediators_mutation_summary, 
        type = 'complete', additionalFilename = "./DMAresults/")
```


```{r, fig.width=6, fig.height=4, echo = FALSE, fig.align="center",hide=TRUE, message=FALSE,warning=FALSE}
img <- readPNG("heatmap_complete.png")
grid.raster(img)
```


## `plotMoonlight`: Heatmap of Moonlight gene z-score for the TSGs/OCGs
In the following plot the top 50 genes with the most driver mutations are visualised.
The values are the moonlight gene z-score for the two biological processes

```{r, eval = FALSE, echo = TRUE, results='hide', warning = FALSE, message = FALSE}
data(DEG_Mutations_Annotations)
data(Oncogenic_mediators_mutation_summary)
data(dataURA)
plotMoonlight(DEG_Mutations_Annotations,
        Oncogenic_mediators_mutation_summary, 
        dataURA, gene_type = "drivers", n = 50)
```

```{r, fig.width=6, fig.height=4, echo = FALSE, fig.align="center",hide=TRUE, message=FALSE,warning=FALSE}
img <- readPNG("moonlight_heatmap.png")
grid.raster(img)
```


# Moonlight Analysis: Case Studies

### Introduction 
This vignette shows a complete workflow of the 'Moonlight2R' package.
The code is divided into four case studies:

* 1. Predicting oncogenic mediators using Moonlight's primary layer 
* 2. Moonlight pipeline in one function
* 3. Moonlight with driver mutation analysis

## Case study n. 1: Predicting oncogenic mediators using Moonlight's primary layer

```{r,eval = FALSE,echo=TRUE,message=FALSE,warning=FALSE, results='hide'}
data(DEGsmatrix)
data(dataFilt)

dataFEA <- FEA(DEGsmatrix = DEGsmatrix)

dataGRN <- GRN(TFs = rownames(DEGsmatrix), 
               DEGsmatrix = DEGsmatrix,
               DiffGenes = TRUE,
               normCounts = dataFilt,
               nGenesPerm = 2000,
               nBoot = 400)

dataURA <- URA(dataGRN = dataGRN, 
              DEGsmatrix = DEGsmatrix, 
              BPname = c("apoptosis",
                         "proliferation of cells"))

dataDual <- PRA(dataURA = dataURA, 
               BPname = c("apoptosis",
                          "proliferation of cells"),
               thres.role = 0)

oncogenic_mediators <- list("TSG"=names(dataDual$TSG), "OCG"=names(dataDual$OCG))

```


## `plotURA`: Upstream regulatory analysis plot
The user can plot the result of the upstream regulatory analysis using the function `plotURA`.

```{r, eval = FALSE,message=FALSE,warning=FALSE, results='hide'}
 plotURA(dataURA = dataURA[c(names(dataDual$TSG), names(dataDual$OCG)),, drop = FALSE], additionalFilename = "_exampleVignette")
```

The figure resulted from the code above is shown below:
```{r, fig.width=6, fig.height=4, echo = FALSE, fig.align="center",hide=TRUE, message=FALSE,warning=FALSE}
img <- readPNG("URAplot.png")
grid.raster(img)
```


## Case study n. 2: Moonlight pipeline in one function

```{r,eval = FALSE,echo=TRUE,message=FALSE,warning=FALSE}
data(dataFilt)
data(DEGsmatrix)
data(dataMAF)
listMoonlight <- moonlight(dataDEGs = DEGsmatrix,
                           dataFilt = dataFilt,
                           nTF = 100,
                           DiffGenes = TRUE,
                           BPname = c("apoptosis","proliferation of cells"),
                           dataMAF = dataMAF,
                           path_cscape_coding = "css_coding.vcf.gz", 
                           path_cscape_noncoding = "css_noncoding.vcf.gz")
save(listMoonlight, file = paste0("listMoonlight_ncancer4.Rdata"))


```

## `plotCircos`: Moonlight Circos Plot
An example of running Moonlight on five cancer types is visualized below in a circos plot.
Outer ring: color by cancer type, Inner ring: OCGs and TSGs, 
Inner connections:  green: common OCGs yellow: common TSGs red: possible dual role

```{r, eval = TRUE, echo = TRUE, results='hide', warning = FALSE, message = FALSE}
data(listMoonlight)
plotCircos(listMoonlight = listMoonlight, additionalFilename = "_ncancer5")
```

The figure generated by the code above is shown below:
```{r, fig.width=6, fig.height=4, echo = FALSE, fig.align="center",hide=TRUE, message=FALSE,warning=FALSE}
img <- readPNG("circos_ocg_tsg_ncancer5.png")
grid.raster(img)
```


## Case study n. 3: Moonlight with driver mutation analysis  

```{r,eval = FALSE,echo=TRUE,message=FALSE,warning=FALSE}

data(DEGsmatrix)
data(dataFilt)
data(dataMAF)


# Perform gene regulatory network analysis
dataGRN <- GRN(TFs = rownames(DEGsmatrix), 
               normCounts = dataFilt,
               nGenesPerm = 2000, 
               kNearest = 3, 
               nBoot = 400)

# Perform upstream regulatory analysis
# As example, we use apoptosis and proliferation of cells as the biological processes
dataURA <- URA(dataGRN = dataGRN,
               DEGsmatrix = DEGsmatrix,
               BPname = c("apoptosis", 
                          "proliferation of cells"), 
               nCores = 1)

# Perform pattern recognition analysis
dataPRA <- PRA(dataURA = dataURA,
               BPname = c("apoptosis", 
                          "proliferation of cells"),
               thres.role = 0)

# Perform driver mutation analysis
dataDMA <- DMA(dataMAF = dataMAF,
               dataDEGs = DEGsmatrix, 
               dataPRA = dataPRA,
               results_folder = "DMAresults",
               coding_file = "css_coding.vcf.gz",
               noncoding_file = "css_noncoding.vcf.gz")

```

Next, we analyze the predicted driver genes and their mutations. 

```{r}
# Extract oncogenic mediators that contain at least one driver mutation
# These are the driver genes
knitr::kable(Oncogenic_mediators_mutation_summary %>%
  filter(!is.na(CScape_Driver)))

# Extract mutation annotations of the predicted driver genes
driver_mut <- DEG_Mutations_Annotations %>% 
  filter(!is.na(Moonlight_Oncogenic_Mediator), 
         CScape_Mut_Class == "Driver")

# Extract driver genes with a predicted effect on the transcriptional level
transcription_mut <- Oncogenic_mediators_mutation_summary %>%
  filter(!is.na(CScape_Driver)) %>%
  filter(Transcription_mut_sum > 0)

# Extract mutation annotations of predicted driver genes that have a driver mutation
# in its promoter region with a potential effect on the transcriptional level 
promoters <- DEG_Mutations_Annotations %>% 
  filter(!is.na(Moonlight_Oncogenic_Mediator), 
         CScape_Mut_Class == "Driver",
         Potential_Effect_on_Transcription == 1,
         Annotation == 'Promoter')
```


## Citation

Please cite the MoonlightR and Moonlight2R packages:

* "Interpreting pathways to discover cancer driver genes with Moonlight." Nature Communications (2020): [10.1038/s41467-019-13803-0](https://doi.org/10.1038/s41467-019-13803-0). [@ref26]

* "An Automatized Workflow to Study Mechanistic Indicators for Driver Gene Prediction with Moonlight." bioRxiv (2022): [10.1101/2022.11.18.517066](https://doi.org/10.1101/2022.11.18.517066). [@ref27]

Related publications to this vignette:

* "TCGAbiolinks: an R/Bioconductor package for integrative analysis of TCGA data." Nucleic acids research (2015): [gkv1507](http://dx.doi.org/doi:10.1093/nar/gkv1507). [@ref25]

* "TCGA Workflow: Analyze cancer genomics and epigenomics data using Bioconductor packages". F1000Research [10.12688/f1000research.8923.1](http://dx.doi.org/doi:10.12688/f1000research.8923.1) [@ref24]


******

Session Information
******
```{r sessionInfo}
sessionInfo()
```

# References
